\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis

### 2023-01-26 src/tex/ms.tex
minor build pdf fix
fixing stuff in sec 3

Derivatives are ubiquitously useful throughout data analysis tasks.
% For instance, during PE, gradients are used to improve performance of the samplers 
% % \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
% (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically required a significant amount of work to obtain~\citep{Keppel:2013kia}.
With waveforms only increasing in complexity, calculating analytic derivatives will become ever more tricky.
% \kw{This needs some rewording. For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms. Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.

@@

\begin{figure}[t]
	\script{loss_compare.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
		\caption{
			Relative error between the numerical relativity and IMRPhenomD waveform amplitudes. 
            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
            Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
		\label{fig:loss_compare}
	\end{centering}
\end{figure}

@@

The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within 20\%.\te{Check number}
This demonstrates that AD can be used to accurately produce population-level forecasts.

@@

\amc{most acronyms were already introduced.}
After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called PE.
To do this, one typically uses a Markov Chain Monte Carlo (MCMC) or nested sampler~\citep{Skilling2004, multinest, dynesty}.
Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$ hours, while BNS systems can take up to weeks.
Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.~\texttt{BAYESTAR}][]{Singer:2015ema}.
\AZ{We should be careful here. Many new algorithms have sped this a lot. Cogwheel can do fast PE, DINGO is a normalizing flow code that is very fast when trained. There are likelihood acceleration methods like ROQ, multibanding, relative binning/heterodyning that accelerate the process compared to the numbers quoted above.}
Nevertheless, these do not present the whole picture; fast, general PE therefore remains a key aim of GW data analysis

A primary issue with both MCMC and nested sampling is that neither utilizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
Derivative based samplers, on the other hand, have been shown to extrapolate well to higher dimensions although they sometimes come with their own drawbacks.
Here we simply aim to demonstrate the utility of a derivative based sampler and their efficiency on a small test problem.
% In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm.
\AZ{Add citations to other HMC papers in GW land}

For our basic example we perform an injection recovery test on a seven dimensional parameter space with the same three detector network setup considered in \S~\ref{subsec:fisher}.
\AZ{As before, say what PSD and cite.}
More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: chirp mass $M_c = 23.82\,\mathrm{M_\odot}$,  symmetric mass ratio $\eta = 0.248$, primary spin parameter $\chi_1=0.3$, secondary spin parameter \AZ{$\chi_2=-0.4$}, luminosity distance $D = 1.6\,\mathrm{Gpc}$, coalescence time $t_c = 0.0$, and coalescence phase $\phi_c = 0.0$.\footnote{
    The remaining parameters are set to $\pi/3$. \AZ{Can we say what they are? $\iota$, $\psi$, $\alpha$, $\delta$?}
}
Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for \sout{200000} $2\times10^5$ steps and the Gaussian random walk (GRW) sampler for \sout{1500000} $1.5\times10^6$ steps (each with four randomly initialized independent chains). 
\AZ{What implementation of GRW? Can we cite something?}
\amc{briefly summarize and give refs for flowMC and GRW. Is GRW Metropolis-Hastings with a Gaussian transition kernel? How'd you tune this? Why not use NUTS for HMC?}
The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
\amc{not a huge deal, but why hand tune? There are automated ways to do this in e.g.\ numpyro with a warm-up adaptation phase.}
The additional steps for the GRW sampler were required to achieve a similarly converged posterior.
\amc{Why not plot its posterior?}

@@

% To illustrate the benefit of a gradient based sampler (such as HMC), we run the same sampling as above but instead use a Gaussian random walk (GRW) sampler. 
% We sample for the same number of steps as used in the HMC with identical initializations to explore acceptance rates of new points.
% For the HMC example, find that the mean local acceptance rate across the 15 chains is around $\sim0.28$, where as the local acceptance rate for the equivalent GRW sampler is XXX.
% Since we are running the chains from the same initialization, this enhanced acceptance for HMC can be mainly associated with use of gradient information to guide sampling.
Although further steps would be required to achieve a fully-converged posterior, these chains are sufficient to show the increased efficiency associated with HMC. 
To further illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and GRW samplers.
The HMC autocorrelation is substantially lower than that of the GRW.
We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
In addition, we found that the effictive number of samples 
\amc{provide ref?} 
(a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.

In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}. 
We therefore expect \jax waveforms to be \sout{crucial} \AZ{highly beneficial} to future PE efforts in GW astronomy, especially for low-latency pipelines.
\kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}

>>>>

Derivatives are ubiquitously useful throughout data analysis tasks.
% For instance, during PE, gradients are used to improve performance of the samplers 
% % \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
% (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\,using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\, using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically required a significant amount of work to obtain~\citep{Keppel:2013kia}.
With waveforms only increasing in complexity, calculating analytic derivatives will become ever more tricky.
% \kw{This needs some rewording. For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms. Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.

@@

\begin{figure}[t]
	\script{loss_compare.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
		\caption{
			Relative error between the numerical relativity and IMRPhenomD waveform amplitudes. 
            The blue line shows the relative error of the IMRPhenomD model with the original model coefficients where as the orange line uses the IMRPhenomD model after optimization using gradient descent.
            Interestingly, the error is reduced across the frequency range, demonstrating that high dimensional optimization is useful for all parts of the waveform.
        }
		\label{fig:loss_compare}
	\end{centering}
\end{figure}

@@

The distribution of sky localization errors from a population of $10^3$ binaries can be seen in Fig.~\ref{fig:sky_localization}.
We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within ~30\%.
This demonstrates that AD can be used to accurately produce population-level forecasts.

@@

% \amc{most acronyms were already introduced.}
After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called PE.
To do this, one typically uses an MCMC or nested sampler~\citep{Skilling2004, multinest, dynesty}.
Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$ hours, while BNS systems can take up to weeks.
Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.~\texttt{BAYESTAR}][]{Singer:2015ema}. 
Moreover, a number of methods have been developed to speed up PE well below the numbers quoted above~\citep{Dax:2021tsq, Islam:2022afg,Roulet:2022kot, Zackay:2018qdy, Cornish:2021lje, Canizares:2013ywa}.
% \AZ{We should be careful here. Many new algorithms have sped this a lot. Cogwheel can do fast PE, DINGO is a normalizing flow code that is very fast when trained. There are likelihood acceleration methods like ROQ, multibanding, relative binning/heterodyning that accelerate the process compared to the numbers quoted above.}
Nevertheless, these do not present the whole picture; fast, general PE therefore remains a key aim of GW data analysis

A primary issue with both MCMC and nested sampling is that neither utilizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
Derivative based samplers, on the other hand, have been shown to extrapolate well to higher dimensions although they sometimes come with their own drawbacks.
Here we simply aim to demonstrate the utility of a derivative based sampler and their efficiency on a small test problem.
% In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
In particular, we will show that the autocorrelation of a Hamiltonian Monte Carlo (HMC) sampler is significantly lower than a traditional MCMC algorithm~\citep{Bouffanais:2018hoz, 2014CQGra..31n5004P}.

For our basic example we perform an injection recovery test on a seven dimensional parameter space with the same three detector network setup considered in \S~\ref{subsec:fisher}.
Again, our noise curves correspond to the design PSDs for LIGO Hanford, LIGO Livingston (\textsc{SimNoisePSDaLIGOZeroDetHighPower}), and Virgo (\textsc{SimNoisePSDAdvVirgo})
We use $20\,$Hz, $1024\,$Hz, and $16\,$s for the minimum frequency, sampling frequency, and sample length.
More specifically, we generate Gaussian noise consistent with the measured PSDs for each detector and then inject a BBH signal with parameters: chirp mass $M_c = 23.82\,\mathrm{M_\odot}$,  symmetric mass ratio $\eta = 0.248$, primary spin parameter $\chi_1=0.3$, secondary spin parameter $\chi_2=-0.4$, luminosity distance $D = 1.6\,\mathrm{Gpc}$, coalescence time $t_c = 0.0$, and coalescence phase $\phi_c = 0.0$.\footnote{
    The remaining parameters (inclination angle $\iota$, polarization angle $\psi$, right ascension $\alpha$, declination $\delta$) are set to $\pi/3$.
}
Using a standard Gaussian likelihood, we then run the HMC sampler implemented in \flowMC for \sout{200000} $2\times10^5$ steps and the Gaussian random walk (GRW) sampler for \sout{1500000} $1.5\times10^6$ steps (each with four randomly initialized independent chains). 
\AZ{What implementation of GRW? Can we cite something?}
\amc{briefly summarize and give refs for flowMC and GRW. Is GRW Metropolis-Hastings with a Gaussian transition kernel? How'd you tune this? Why not use NUTS for HMC?}
The number of steps and and mass matrix used for each example was hand tuned to give good performance for the specific sampler.
\amc{not a huge deal, but why hand tune? There are automated ways to do this in e.g.\ numpyro with a warm-up adaptation phase.}
The additional steps for the GRW sampler were required to achieve a similarly converged posterior.
\amc{Why not plot its posterior?}

@@

% To illustrate the benefit of a gradient based sampler (such as HMC), we run the same sampling as above but instead use a Gaussian random walk (GRW) sampler. 
% We sample for the same number of steps as used in the HMC with identical initializations to explore acceptance rates of new points.
% For the HMC example, find that the mean local acceptance rate across the 15 chains is around $\sim0.28$, where as the local acceptance rate for the equivalent GRW sampler is XXX.
% Since we are running the chains from the same initialization, this enhanced acceptance for HMC can be mainly associated with use of gradient information to guide sampling.
Although further steps would be required to achieve a fully-converged posterior, these chains are sufficient to show the increased efficiency associated with HMC. 
To further illustrate this, in Fig.~\ref{fig:autocorrelation} we plot the autocorrelation as a function of lag for both the HMC and GRW samplers.
The HMC autocorrelation is substantially lower than that of the GRW.
We therefore expect gradient based samplers to converge significantly faster than typical samplers, especially in higher dimensions.
In addition, we found that the effictive number of samples~\citep{arviz_2019},\footnote{
    Computed using \textsc{arviz}~\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}
    }
(a measure of the number of indepedent samples) is between 2 and 7 times larger for HMC across the different dimensions of the parameter space.

In a follow up paper we will demonstrate that minute scale PE can be achieved by combining normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu}, GPU acceleration, and a derivative based sampler~\citep{PEpaper}. 
We therefore expect \jax waveforms to be highly beneficial to future PE efforts in GW astronomy, particular for low-latency pipelines and higher dimensional analysese.
% \kw{Not just low latency, but also high dimensional analysis, such as testing GR and lensing.}
